---
title: "510_Final_Project"
author: "Hanna Lu"
date: "12/11/2020"
output: html_document
---


### Survival Analysis

The aim of this study is to find significant factors that impact the mortality rate of critically ill patients within 90 days after they are discharged from the hospital. The findings of this study may help doctors to re-evaluate their discharge decisions to reduce the mortality rate.

This analysis uses variables from multiple data sets. Some of them are re-engineered. In addition to demographic and hospital stay information, comorbidity scores are also investigated in this study.

#### Data Source: https://mimic.physionet.org/gettingstarted/access/

#### Responses and updates to Peer Review comments::
* 1. The data is de-identified. The date of birth and date of death (if applicable) are masked. There's no other information that can identify the patient. But nevertheless, the data is restricted.
* 2. The inconsistency in subsetting the data by columns has been fixed.
* 3. About training and test sets, since this project is to identify factors that impact the mortality and to analyze the effect of the impact, it is more explanatory than predictive at this point, and thus no need to split the data into training and test sets. However, our future work will include a prediction part after investigating more variables such as vital sign data and topic modeling results from doctor’s notes, which will require splitting the data into training and test sets.
* 4. Added a future work section at the end.

```{r setup, echo=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/520ha/Desktop/Chapman/510_Survival_Analysis")
```

Install packages:
```{r message=FALSE, warning=FALSE}

library(dplyr)
library(stringr)
library(reshape2)
library(comorbidity)
library(ggplot2)
library(survival)
library(survminer)
library(grid)
library(tiff)
```

### IMPORT NOTEEVENTS
<br/>
A patient can have multiple reports (or doctor's notes). We hypothesize that the more doctor's notes, the higher the risk of mortality.But the number of reports is not readily available in the data set.

The following code creates a variable NUM_REPORTS (Number of Reports) from the NOTEEVENTS data set by grouping the data by HADM_ID (Hospital Admission ID) and then counting the number of rows in each group.
```{r}
filename <- gzfile('Data/NOTEEVENTS.csv.gz','rt')
note <- read.csv(filename)

str(note)

# Create a data frame with 2 columns (HADM_ID and NUM_REPORTS) to be joined
# with other data sets
num_reports <- note %>%
  group_by(HADM_ID) %>%
  summarise(NUM_REPORTS = n(),.groups = 'drop') %>%
  select(HADM_ID, NUM_REPORTS)
```

### IMPORT ADMISSIONS

The ADMISSIONS data set contains some demographic information of the patients, as well as hospital admission and discharge information 

The following code joins this data set with the Number of Reports data extracted from the NOTEEVENTS data set.

```{r}
filename <- gzfile('Data/ADMISSIONS.csv.gz','rt')
admissions <- read.csv(filename)

str(admissions)

# Merge admissions and num_reports (inner join):
adm = merge(admissions, num_reports, by='HADM_ID')
```

### IMPORT ICUSTAYS

This data set contains the ICU stay information such as the first and last ICU unit the patient stayed in, in and out time, and length of stay.

A patient can be admitted to different ICU units during their stay in the hospital. We hypothesize that the last ICU unit the patient stayed in may be related to the patient's risk of mortality since it carries the information about the patient's latest condition. ICU length of stay may have an impact on the mortality risk as well.

The following code extracts the last ICU unit the patient stayed in, and the total length of stay in all ICU units.

```{r warning=FALSE}
filename <- gzfile('Data/ICUSTAYS.csv.gz','rt')
icu <- read.csv(filename)

str(icu)

# Create two columns for the last ICU unit and total Length of Stay:
icu_last <- icu %>%
  select(HADM_ID, LAST_CAREUNIT, LOS) %>%
  group_by(HADM_ID) %>%
  mutate(LAST_ICU = last(LAST_CAREUNIT)) %>%
  select(HADM_ID, LAST_ICU) %>%
  distinct()

# Merge ICU Unit data with admission data (inner join):
adm_icu <- merge(adm, icu_last, by = 'HADM_ID')
```


### IMPORT PATIENTS

This data set contains patients' personal information such as gender, date of birth, and date of death. We need the date of birth variable to calculate the patients' age, and the date of death variable to create the EVENT column (weather of not a death event occurred) and the Duration column (length of time until a death event occurred) for the survival analysis.

The following code joins this data set with the previous data set.

```{r}
filename <- gzfile('Data/PATIENTS.csv.gz','rt')
patients <- read.csv(filename)

str(patients)

# Merge patients and adm_icu:
df_patient_icu <- merge(adm_icu, patients, by='SUBJECT_ID', all.x = T)
```

### IMPORT DIAGNOSES_ICD

This data set contains the ICD9_CODE (code for type of diagnosed disease) for each patient. A patient can have multiple ICD9 codes

```{r warning=FALSE, message=FALSE}
filename <- gzfile('Data/DIAGNOSES_ICD.csv.gz','rt')
diagnoses <- read.csv(filename)

diagnoses <- diagnoses[,c("HADM_ID","ICD9_CODE")]

df_com <- merge(df_patient_icu, diagnoses, by = 'HADM_ID', all.x = T)
```


### COMORBIDITY SCORES
The following code calculates the comorbidity scores using the comorbidity package. The comorbidity function generates 17 comorbidity scores, of which, "score", "index","wscore",and "windex" are overall scores, and the rest are specific to a certain disease. Therefore, the 4 overall scores are correlated with the specific scores, and thus need to be removed.

```{r warning=FALSE, message=FALSE}
# Charlson comorbidity scores based on ICD-10 diagnostic codes:
df_comorbidity <- comorbidity(x = df_com, id = "HADM_ID", code = "ICD9_CODE", 
                             score = "charlson", assign0 = FALSE, icd = 'icd9')

# Remove the last four items from the comorbidity output
df_comorbidity <- df_comorbidity[,!names(df_comorbidity) %in% 
                                   c("score", "index","wscore","windex")]

# Remove the icd code column from original data frame:
com_no_icd <- df_com[,names(df_com) != 'ICD9_CODE']

# Remove duplicates from original data frame:
unique_com <- unique(com_no_icd)

# Merge the main dataset and the comorbidity:
com_all <- merge(unique_com, df_comorbidity, by="HADM_ID")

# Remove "DIAGNOSIS":
com_all <- select(com_all, -DIAGNOSIS)
```

### DATA CLEANING

The following code generates the EVENT column (whether or not a patient died within 90 days of discharge from hospital) and the Duration column (time difference between Date of Death and Date of Discharge if the patient died during the following up period. If Date of Death is missing, set it as 300 days)

The following code also removes the patients who are admitted to the hospital as "NEWBORN", or died within one day after discharge, or died in the hospital.

```{r}
# Convert DOD (Date of Death) to date format:
com_all$DOD <- as.Date(com_all$DOD, format = '%Y-%m-%d')

df_all <- com_all %>%
  # Identify DODs after 90 days and create Duration column
  mutate(Duration = ifelse(is.na(DOD), 300, 
                           (DOD - as.Date(DISCHTIME, format = '%Y-%m-%d %H:%M:%S')))) %>%
  # Create EVENT (death within 90 days of discharge as 1, otherwise 0)
  mutate(EVENT = ifelse(Duration <= 90, 1, 0)) %>%
  # Remove patients who died before discharge or within one day of discharge
  # Remove newborns
  # Remove patients who died in hospital
  filter(Duration >= 1 & ADMISSION_TYPE!='NEWBORN' & HOSPITAL_EXPIRE_FLAG == 0)
```

The following code creates the HOSPITAL_LOS (Hospital Length of Stay, which is time between admission and discharge) and Age(age at time of discharge)  columns. It also removes the patients who stayed in the hospital for less than one day, who are under the age of 18, and those who are discharged to hospice.

```{r}
df_all <- df_all %>%
  # Create a column for hospital length of stay
  mutate(HOSPITAL_LOS = (as.Date(DISCHTIME, format = '%Y-%m-%d %H:%M:%S') - 
                           as.Date(ADMITTIME, format = '%Y-%m-%d %H:%M:%S'))) %>%
  # Remove those who stayed less than 1 day in the hospital
  filter(HOSPITAL_LOS > 1) %>%
  # Create Age at Discharge column from DOB and Discharge time columns
  mutate(age = (as.Date(DISCHTIME, format = '%Y-%m-%d %H:%M:%S') -
                  as.Date(DOB, format = '%Y-%m-%d %H:%M:%S'))/365) %>%
  # Remove those who are under the age of 18:
  filter(age >= 18) %>%
  # Remove those that were discharged to hospice related locations
  filter(DISCHARGE_LOCATION != 'HOSPICE')

# If age is greater than 100, code it to 90
# In the MIMIC database, patients older than 89 were coded as 300 years old
df_all$age <- ifelse(df_all$age > 100, 90, round(df_all$age,1))
```

The following code regroups categorical variables, and removes patients who were sent to hospice.

```{r}
# Regroup Marital Status:
df_all <- df_all %>% 
  mutate(MARITAL_STATUS_CAT = case_when(
          MARITAL_STATUS == 'SINGLE'                       ~ 'SINGLE',
          MARITAL_STATUS == 'WIDOWED'                      ~ 'WIDOWED',
          MARITAL_STATUS %in% c('MARRIED', 'LIFE PARTNER') ~ 'MARRIED',
          MARITAL_STATUS %in% c('DIVORCED','SEPARATED')    ~ 'DIVORCED',
          TRUE                                             ~ 'UNKNOWN'))

df_all$MARITAL_STATUS_CAT <- factor(df_all$MARITAL_STATUS_CAT,
                                    levels = c("MARRIED","SINGLE","DIVORCED",
                                       "WIDOWED", "UNKNOWN"))

# Regroup Ethnicity:
df_all <- df_all %>%
  mutate(ETHNICITY_CAT = case_when(
    str_detect(ETHNICITY, 'WHITE')                               ~ 'WHITE',
    str_detect(ETHNICITY, 'BLACK')                               ~ 'BLACK',
    str_detect(ETHNICITY, 'HISPANIC')                            ~ 'HISPANIC',
    str_detect(ETHNICITY, 'ASIAN')                               ~ 'ASIAN',
    ETHNICITY %in% c('UNKNOWN/NOT SPECIFIED','UNABLE TO OBTAIN',
                     'PATIENT DECLINED TO ANSWER')               ~ 'UNKNOWN',
    TRUE                                                         ~ 'OTHER'))

df_all$ETHNICITY_CAT <- factor(df_all$ETHNICITY_CAT, 
                               levels = c("WHITE","BLACK","HISPANIC","ASIAN",
                                          "OTHER","UNKNOWN"))

# Discharge Location:
df_all <- df_all %>%
  mutate(DISCH_LOC_CAT = case_when(
    str_detect(DISCHARGE_LOCATION, 'SNF')            ~ 'SNF',
    DISCHARGE_LOCATION %in% c('HOME HEALTH CARE',
                      'HOME WITH HOME IV PROVIDER')  ~ 'HOME HEALTH CARE',
    DISCHARGE_LOCATION == 'HOME'                     ~ 'HOME',
    DISCHARGE_LOCATION == 'LONG TERM CARE HOSPITAL'  ~ 'LONG TERM CARE HOSPITAL',
    DISCHARGE_LOCATION == 'REHAB/DISTINCT PART HOSP' ~ 'REHAB/DISTINCT PART HOSP',
    TRUE                                             ~ 'OTHER')) 

df_all$DISCH_LOC_CAT <- factor(df_all$DISCH_LOC_CAT, 
                               levels = c("HOME","HOME HEALTH CARE",
                                          "LONG TERM CARE HOSPITAL",
                                          "REHAB/DISTINCT PART HOSP",
                                          "SNF","OTHER"))
```

The following code makes numerical variables into categorical.

```{r}
# Impute the NUM_REPORTS column with 0 where Num_reports is missing
df_all$NUM_REPORTS <- ifelse(is.na(df_all$NUM_REPORTS), 0, df_all$NUM_REPORTS)

# Make numerical variables into bins:
df_all <- within(df_all,{
  # Bin Hospital Length of Stay:
  HOSPITAL_LOS_CAT <- cut(as.numeric(HOSPITAL_LOS), breaks=c(-Inf,5,10,20,Inf),
                         labels = c("<=5","6-10","11-20",">20"))
  # Bin Number of Reports:
  NUM_REPORTS_CAT <- cut(NUM_REPORTS, breaks=c(-Inf,10,20,50, Inf),
                        labels = c("<=10","11-20", "21-50", ">50"))
  # Bin Age according to MeSH:
  AGE_CAT <- cut(round(age), breaks=c(-Inf,44,64,79, Inf),
                labels = c("<=44","45-64", "65-79", ">=80"))
})

# Reorder the levels of the new categorical variables:
df_all$HOSPITAL_LOS_CAT = factor(df_all$HOSPITAL_LOS_CAT,
                                 levels = c("<=5","6-10","11-20",">20"))

df_all$NUM_REPORTS_CAT = factor(df_all$NUM_REPORTS_CAT, 
                                levels = c("<=10","11-20", "21-50", ">50" ))

df_all$AGE_CAT = factor(df_all$AGE_CAT, 
                        levels = c("<=44", "45-64", "65-79", ">=80"))

```


### DATA VISUALIZATION

```{r fig1, fig.height = 3, fig.width = 5, fig.align = "center", message=FALSE, warning=FALSE}
# Plot of Hospital Length of Stay vs Event
ggplot(df_all) + 
  geom_bar(aes(x = HOSPITAL_LOS_CAT, fill = factor(EVENT)),width = 0.5) +
  ggtitle('Hospital Length of Stay by Event') + 
  theme(plot.title = element_text(hjust = 0.5, size = 12)) +
  labs(x = 'Hospital Length of Stay (days)', y = 'Number of Patients', 
       fill = 'Event')
```

```{r fig2, fig.height = 3, fig.width = 5, fig.align = "center", message=FALSE, warning=FALSE}
# Plot of Number of Reports by Event
ggplot(df_all) + 
  geom_bar(aes(x = NUM_REPORTS_CAT, fill = factor(EVENT)), width = 0.5) +
  ggtitle('Number of Reports by Event') + 
  theme(plot.title = element_text(hjust = 0.5, size = 12)) +
  scale_color_discrete(name = 'Event') +
  labs(x = 'Number of Reports', y = 'Number of Patients', fill = 'Event')
```

```{r fig3, fig.height = 3, fig.width = 5, fig.align = "center", message=FALSE, warning=FALSE}
# Plot of Age by Event
ggplot(df_all) + 
  geom_bar(aes(x = AGE_CAT, fill = factor(EVENT)), width = 0.5) +
  ggtitle('Age by Event') + 
  theme(plot.title = element_text(hjust = 0.5, size = 12)) +
  scale_color_discrete(name = 'Event') +
  labs(x = 'Age Group', y = 'Number of Patients', fill = 'Event')
```

```{r fig4, fig.height = 3, fig.width = 5, fig.align = "center", message=FALSE, warning=FALSE}
# Plot of Admission Type by Event
ggplot(df_all) + 
  geom_bar(aes(x = ADMISSION_TYPE, fill = factor(EVENT)), width = 0.5) +
  ggtitle('Admission Type by Event') + 
  theme(plot.title = element_text(hjust = 0.5, size = 12)) +
  scale_color_discrete(name = 'Event') +
  labs(x = 'Admission Type', y = 'Number of Patients', fill = 'Event')
```

```{r fig5, fig.height = 3, fig.width = 5, fig.align = "center", message=FALSE, warning=FALSE}
# Plot of MARITAL_STATUS_CAT by Event
ggplot(df_all) + 
  geom_bar(aes(x = MARITAL_STATUS_CAT, fill = factor(EVENT)), width = 0.5) +
  ggtitle('Marital Status by Event') + 
  theme(plot.title = element_text(hjust = 0.5, size = 12),
        axis.text.x = element_text(angle = 45, vjust=1, hjust=1)) +
  scale_color_discrete(name = 'Event') +
  labs(x = 'Marital Status', y = 'Number of Patients', fill = 'Event')
```

```{r fig6, fig.height = 3, fig.width = 5, fig.align = "center", message=FALSE, warning=FALSE}
# Plot of ETHNICITY_CAT by Event
ggplot(df_all) + 
  geom_bar(aes(x = ETHNICITY_CAT, fill = factor(EVENT)), width = 0.5) +
  ggtitle('Ethnicity by Event') + 
  theme(plot.title = element_text(hjust = 0.5, size = 12),
        axis.text.x = element_text(angle = 45, vjust=1, hjust=1)) +
  scale_color_discrete(name = 'Event') +
  labs(x = 'Ethnicity', y = 'Number of Patients', fill = 'Event')
```


### SURVIVAL ANALYSIS

```{r}

# Select columns that will be used in the model:
all <- df_all %>%
  select("HADM_ID","ADMISSION_TYPE","LAST_ICU", "GENDER",
         "ami","chf","pvd","cevd","dementia","copd","rheumd","pud","mld",          
         "diab","diabwc" ,"hp", "rend","canc","msld","metacanc","aids",
         "Duration", "EVENT","MARITAL_STATUS_CAT", "ETHNICITY_CAT",    
         "AGE_CAT","NUM_REPORTS_CAT","HOSPITAL_LOS_CAT")
```

<p>&nbsp;</p>
The following code performs the Cox Proportional Hazard survival analysis:

```{r message=FALSE, warning=FALSE}
time <- all$Duration
event <- all$EVENT
x <- subset(all, select = -c(HADM_ID, Duration, EVENT))

coxph <- coxph(Surv(time,event) ~ ., data = x)
summary(coxph)
```

<p>&nbsp;</p>
The following code saves the summary of the output of the coxph model with all variables to the results folder.
```{r warning=FALSE, message=FALSE}
# Save model output:
sink(file = 'Results/cox_all_vars.txt')
summary(coxph)
sink(file=NULL)
```

<p>&nbsp;</p>
The following code removes all insignificant variables and then reruns the model.
```{r}
x <- select(x, -c(diab,pud,rheumd,diabwc))
                  
coxph <- coxph(Surv(time,event) ~ ., data = x)
summary(coxph)
```

<p>&nbsp;</p>
The following code saves the summary of the output of the coxph model with significant variables only to the results folder.
```{r warning=FALSE, message=FALSE}
# Save model output:
sink(file = 'Results/cox_sig_vars.txt')
summary(coxph)
sink(file=NULL)
```

<p>&nbsp;</p>
The following code plots the hazard ratio based on the final model with significant variables only, and saves it to the Results folder. 

In the plot, the intervals that are completely on the right side of the dashed line suggest that the corresponding variables significantly contribute to mortality; the intervals that are completely on the left side of the dashed line suggest that the corresponding variables reduce mortality; and the intervals that cross the dash line suggest that the corresponding variables are not significant.

```{r fig7, fig.height = 18, fig.width = 10, fig.align = "center", message=FALSE, warning=FALSE}
# Plot the hazard ratio
ggforest(coxph, data = x, main="Hazard Ratio", fontsize = 0.7)
```

```{r warning=FALSE, message=FALSE}
# Save the plot:
tiff(file="Results/plot_hazard_ratio.tiff",
     width=10, height=15, units="in", res=300)
ggforest(coxph, data = x, main="Hazard Ratio", fontsize = 0.7)
dev.off()
```


### FINDINGS
  The coef column shows the logarithm of the odds ratio of mortality, and the exp(coef) column shows the odds ratio. A positive coef (or exp(coef) > 1) suggests that the factor is contributing to the mortality, and the larger the absolute value of the coef, the higher the risk of mortality.
  Based on the results from the coxph model with significant factors only, we have the following major findings:
  <p>&nbsp;</p>
  * The concordance score (equivalent to the R2 score in linear regression) is 0.795.
  * The factor that impacts the mortality rate the most is age. The older the patient, the higher the risk of mortality.
  * The second strongest factor that impacts the mortality rate is comorbidity of metacanc, which represents metastatic solid tumor. Patients with metastatic solid tumor have a higher risk of mortality.
  * Patients with msld (moderate or severe liver disease) also have a higher risk of mortality.
  * Patients admitted through the emergency room have a higher risk of mortality than those admitted through urgent care or other ways.
  * The longer the hospital length of stay, the higher the risk of mortality.
  * Marital status and ethnicity are significant, but their impact on mortality is relatively minor.
 
<p>&nbsp;</p>  

### Future Work
  * More variables available in the MIMICIII database will be investigated including certain vital signs such as oxygen saturation, body temperature, heart rate, systolic and diastolic blood pressure.
  * Using topic modeling to analyze the doctor's notes, and incorporate the results in the survival analysis model.
  * Using the significant variables to predict mortality risk.
  
<p>&nbsp;</p>  
